---
title: "2 avoid common statistical problems"
date: "2015-11-16 21:00"
description: explortive data analysis
output: pdf_document
layout: post
draft: yes
published: no
status: process
subtitle: an explorative analysis of soil moisture data
summary: explorative data analysis based on Zuur et al. (2010)
tags: statistics
category: R
---

###Introduction

Alain Zuur, Ieno Elena and Chris Elphick published **A protocol for data exploration to avoid comon statistical problems** in 2010 (Methods in  Ecology and Evolution). They proposed 7 steps in order to prepare a data set towards the use of statistical methods in order to avoid type I or type II errors, thereby reducing the chance of making wrong ecological conclusions or recommendations.

The recomended protocol includes the following questions:

1. Are there outlier in x and y?
2. Do we have homogeneity of variance?
3. Are the data normally distributed?
4. Are there a lot of zeros in the data?
5. Is there collinarity among the covariates?
6. What are the relationships between y and x variables?
7. Should we consider interaction?
8. Are observations of the response variable independent?

In this document we will answer them by means of a visual data exploration.

The data set we are working on is explain in detail within the [R-packages SoilMoisturePattern](). For more information install the package and read the data set documentation:

``` {r, echo = FALSE, include = FALSE}

library(SoilMoisturePattern)

?HiResAlp_MobileCampaigns
data("HiResAlp_MobileCampaigns")

```

###ARE THERE OUTLIER IN X AND Y?

Outlier can be detected with boxplots or Cleveland dotplots (row number vs. observation). 

``` {r, boxplot of HiResALp Muntatschinig campaigns, echo=FALSE, warning=FALSE}

library(ggplot2)

data <- HiResAlp_MobileCampaigns[HiResAlp_MobileCampaigns$CampaignType=="RADAR_MUNTATSCHINIG",]

p_boxes <- 
  ggplot(data = data, aes(y=SoilMoisture_mean_Processed, x=date, group=date)) + 
  geom_hline(aes(linetype="twodash"), yintercept=quantile(data$SoilMoisture_mean_Processed, 
                                                          probs = c(.25,0.5,0.75), na.rm=T), 
             col=c("grey30","grey10","grey30")) + 
  geom_jitter(col="grey") + 
  geom_boxplot(width=0.2, position=position_dodge(width=0.2)) + 
  geom_boxplot(aes(y=precedingRain3_P2), width=0.5, position=position_dodge(width=0.2), col="darkblue") +
  geom_text(aes(x=2,y=65,label="__  preceding Rain (3 days) in P2 [mm]"), col="darkblue", cex=2.5) +
  ylab("smc [vol%]") + 
  xlab("")

p_boxes

```

```{r, Cleveland dotplot of HiResAlp Muntatschinig campaigns, echo=F}

data <- HiResAlp_MobileCampaigns[HiResAlp_MobileCampaigns$CampaignType=="RADAR_MUNTATSCHINIG",]

data <- data[order(data$SoilMoisture_mean_Processed),]
data$rownr <- 1:dim(data)[1]

p_cleveland <-
  ggplot(data = data, aes(x=SoilMoisture_mean_Processed, y=rownr)) +
  geom_point()

p_cleveland

```


Zuur Alain F., Ieno Elena N., Elphick Chris E. (2010): A protocol for data exploration to avoid common statical problems, [Methods in Ecology and Evolution, 1, 3-14](DOI).
